---
title: "COVID-19 Trends"
format: markdown_github
---

This is a guided project for dataquest, investigating COVID-19 trends.

## 2. Understanding the data

```{r}
library(readr)
library(dplyr)
```

Here I'm reading the csv directly from the URL provided; you can download the
file and read it locally as well.

I specify `show_col_types = FALSE` to suppress the verbose messages showing
all of the column types it guessed. It's often useful to see this information
at least the first time you read in a dataset to confirm it is treating the 
columns as you would expect.

```{r}
#| cache: true
covid_df <- read_csv(
  "covid19.csv",
  show_col_types = FALSE
)

getwd()
```

Here we use `dim()` to show the number of rows and columns, then `colnames()`
to get the column names. They are stored in the object called `vector_cols`. By 
running the object `vector_cols` on its own it prints the contents of that 
object (the column names).

```{r}
dim(covid_df)
vector_cols <- colnames(covid_df)
vector_cols
```

`vector_cols` represents a *character vector*

```{r}
head(covid_df)

glimpse(covid_df)
```

`head()` shows you the first `n` rows (where `n` = 6 by default, but you can 
change it).

`glimpse()` gives you a preview of all of the columns and their types

## 3. Isolating the rows we need

Here we use `filter()` to only keep the rows where the value in the 
`Province_State` is equal to `"All States"`. Then we use `select()` and
specify `-Province_state` - the `-` sign *excludes* that column and keeps the rest.

```{r}
covid_df_all_states <- covid_df |>
  filter(Province_State == "All States") |>
  select(-Province_State)

dim(covid_df_all_states)
glimpse(covid_df_all_states)
```

## 4. Isolating the columns we need

Here we use `select()` again, but this time specify the columns we want to keep.

```{r}
covid_df_all_states_daily <- covid_df_all_states |>
  select(
    Date,
    Country_Region,
    active,
    hospitalizedCurr,
    daily_tested,
    daily_positive
  )
```

## 5. Extracting the Top Ten Countries with Most Covid-19 Cases

Using `group_by()` + `summarize()` is a very powerful and common way to 
aggregate values for different levels of a variable. Here we `group_by(Country_Region)`
and then for each value (level) of `Country_Region` we calculate the sum of
several numeric variables.

Then we sort using `arrange()`, and specify `desc(tested)` to sort by `tested` 
in descending order. This can also be done with `arrange(-tested)`.

```{r}
covid_df_all_states_daily_sum <- covid_df_all_states_daily |>
  group_by(Country_Region) |>
  summarise(
    tested = sum(daily_tested),
    positive = sum(daily_positive),
    active = sum(active),
    hospitalized = sum(hospitalizedCurr)
  ) |>
  arrange(desc(tested))

covid_df_all_states_daily_sum
```

We can then use `head()` with `n = 10` on the sorted data frame to get the top
10.

```{r}
covid_top_10 <- head(covid_df_all_states_daily_sum, 10)
covid_top_10
```

## 6. Identifying the Highest Positive Against Tested Cases

Extract the different columns into standalone vectors

```{r}
countries <- covid_top_10$Country_Region
tested_cases <- covid_top_10$tested
positive_cases <- covid_top_10$positive
active_cases <- covid_top_10$active
hospitalized_cases <- covid_top_10$hospitalized
```

Assign the vectors of values (cases) the names of the countries, using the
`countries` character vector.

```{r}
names(tested_cases) <- countries
names(positive_cases) <- countries
names(active_cases) <- countries
names(hospitalized_cases) <- countries

hospitalized_cases
```

Divide positive cases by tested to get positivity rates - look at these manually,
and subset the two vectors by those country names

```{r}
all_postivity <- positive_cases / tested_cases

positive_tested_top_3 <- positive_cases[c("United Kingdom", "United States","Turkey")] /
  tested_cases[c("United Kingdom", "United States", "Turkey")]

positive_tested_top_3
```

### 6a. An alternative using vectors

```{r}
positivity <- sort(positive_cases / tested_cases, decreasing = TRUE)

positive_tested_top_3 <- positivity[1:3]
positive_tested_top_3
```

### 6b. An alternative using data frames

The exercise is about testing your understanding of using and manipulating 
vectors, so the answers above do that well. However if I was doing this in real
life I would keep the data frames and use dplyr to find what I want:

```{r}
covid_top_10 |>
  mutate(positivity = positive / tested) |>
  arrange(desc(positivity)) |>
  slice_head(n = 3)

## OR

covid_top_10 |>
  mutate(positivity = positive / tested) |>
  slice_max(positivity, n = 3)
```

## 7. Keeping relevant information

```{r}
united_kingdom <- c(0.11, 1473672, 166909, 0, 0)
united_states <- c(0.10, 17282363, 1877179, 0, 0)
turkey <- c(0.08, 2031192, 163941, 2980960, 0)
```


```{r}
covid_mat <- rbind(united_kingdom, united_states, turkey)

colnames(covid_mat) <- c(
  "Ratio",
  "tested",
  "positive",
  "active",
  "hospitalized"
)

covid_mat
```

## 8. Putting it all together: Lists

```{r}
question <- "Which countries have had the highest number of positive cases against the number of tests?"
answer <- c("Positive tested cases" = positive_tested_top_3)

question
answer
```

*Today I learned that assigning a name to an already named vector using `c()`
concatenates the new name with the old names, separated by a `.`!*

```{r}
data_structure_list <- list(
  dataframes = list(
    covid_df,
    covid_df_all_states,
    covid_df_all_states_daily,
    covid_top_10
  ),
  matrices = list(covid_mat),
  vectors = list(vector_cols, countries)
)
```

```{r}
covid_analysis_list <- list(
  question,
  answer,
  data_structure_list
)

covid_analysis_list[[2]]
```
