---
title: "Guided Project: Analyzing Forest Fire Data"
format: gfm
---

```{r}
#| include: false

data_file <- "forestfires.csv"

if (!file.exists(data_file)) {
  download.file(
    "https://dsserver-prod-resources-1.s3.amazonaws.com/277/forestfires.csv",
    destfile = data_file
  )
}
```



### 2. The Importance of Forest Fire Data

```{r}
library(tidyverse)

forestfires <- read_csv(
  "forestfires.csv",
  show_col_types = FALSE
)

glimpse(forestfires)
```

Each row represents a single forest fire, with the following columns:

- **X, Y**: Spatial coordinates within the Montesinho park map (1-9 scale)
- **month**: Month of the year (jan, feb, mar, ..., dec)
- **day**: Day of the week (mon, tue, wed, thu, fri, sat, sun)
- **FFMC**: Fine Fuel Moisture Code - measures moisture content of litter and fine fuels. Higher values indicate drier conditions and higher fire risk (0-101 scale)
- **DMC**: Duff Moisture Code - measures moisture content of loosely compacted decomposing matter. Higher values indicate drier conditions in deeper organic layers
- **DC**: Drought Code - measures moisture content in deep, compact organic layers. Indicates seasonal drought effect on forest fuels with slow drying and wetting
- **ISI**: Initial Spread Index - combines wind speed and FFMC to indicate expected rate of fire spread. Higher values indicate faster potential fire spread
- **temp**: Outside temperature in Celsius degrees
- **RH**: Outside relative humidity in percentage
- **wind**: Outside wind speed in km/h
- **rain**: Outside rain in mm/m²
- **area**: Burned area of the forest in hectares. Values of 0 indicate fires that burned less than 1 hectare

The Fire Weather Index System (FFMC, DMC, DC, ISI) uses meteorological inputs to assess fire danger. FFMC responds quickly to weather changes, DMC reflects medium-term drought effects, and DC indicates long-term drought conditions.

## 3: Data Processing

For the months, we can use the built-in `month.abb` object to ensure they are ordered correctly, but we have to convert it to lowercase since our data is all lowercase. For the days of the week, we can use a custom order from Sunday to Saturday.

```{r}
forestfires <- forestfires |>
  mutate(
    month = factor(month, levels = tolower(month.abb), ordered = TRUE),
    day = factor(
      day,
      levels = c("sun", "mon", "tue", "wed", "thu", "fri", "sat"),
      ordered = TRUE
    )
  )

levels(forestfires$month)
levels(forestfires$day)
```

## 4: When Do Most Forest Fires Occur?


```{r}
fires_by_month <- forestfires |>
  group_by(month) |>
  summarise(total_fires = n(), .groups = "drop") |>
  arrange(month)

fires_by_day <- forestfires |>
  group_by(day) |>
  summarise(total_fires = n(), .groups = "drop") |>
  arrange(day)
```


```{r}
ggplot(fires_by_month, aes(x = month, y = total_fires)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Total Forest Fires by Month",
    x = "Month",
    y = "Total Fires"
  ) +
  theme_minimal()
```

```{r}
ggplot(fires_by_day, aes(x = day, y = total_fires)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Total Forest Fires by Day of the Week",
    x = "Day of the Week",
    y = "Total Fires"
  ) +
  theme_minimal()
```

## 5: Plotting Other Variables Against Time


```{r}
forestfires_long <- forestfires |>
  pivot_longer(
    cols = c(FFMC, DMC, DC, ISI, temp, RH, wind, rain),
    names_to = "variable",
    values_to = "value"
  )

forestfires_month_long <- forestfires_long |>
  group_by(month, variable) |>
  summarise(
    avg_value = mean(value, na.rm = TRUE),
    .groups = "drop"
  )
```



```{r}
ggplot(forestfires_month_long, aes(x = month, y = avg_value, fill = variable)) +
  geom_col() +
  labs(
    title = "Average Values of Meteorological Variables by Month",
    x = "Month",
    y = "Average Value"
  ) +
  theme_minimal() +
  # Adjust the x-axis text for better readability
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  # remove the legend since we are using facets that give the same information
  scale_fill_discrete(guide = "none") +
  facet_wrap(~variable, scales = "free_y")
```

## 6: Examining Forest Fire Severity


```{r}
forestfires_long |>
  ggplot(aes(x = value, y = area)) +
  # Using alpha to reduce overplotting
  geom_point(aes(color = variable), alpha = 0.5) +
  facet_wrap(~variable, scales = "free_x") +
  theme_minimal() +
  labs(
    title = "Relationship Between Meteorological Variables and Burned Area",
    x = "Meteorological Variable Value",
    y = "Burned Area (hectares)"
  ) +
  scale_color_discrete(guide = "none")
```

## 7: Outlier Problems

### Remove zero-area fires

```{r}
forestfires_long |>
  filter(area > 0) |>
  ggplot(aes(x = value, y = area)) +
  geom_point(aes(color = variable), alpha = 0.5) +
  facet_wrap(~variable, scales = "free_x") +
  theme_minimal() +
  labs(
    title = "Relationship Between Meteorological Variables and Burned Area",
    x = "Meteorological Variable Value",
    y = "Burned Area (hectares)"
  ) +
  scale_color_discrete(guide = "none")
```

### Remove very large fires

```{r}
forestfires_long |>
  filter(area < 100) |>
  ggplot(aes(x = value, y = area)) +
  geom_point(aes(color = variable), alpha = 0.5) +
  facet_wrap(~variable, scales = "free_x") +
  theme_minimal() +
  labs(
    title = "Relationship Between Meteorological Variables and Burned Area",
    x = "Meteorological Variable Value",
    y = "Burned Area (hectares)"
  ) +
  scale_color_discrete(guide = "none")
```

### Fires between 10 and 200 ha

```{r}
forestfires_long |>
  filter(area >= 10, area <= 200) |>
  ggplot(aes(x = value, y = area)) +
  geom_point(aes(color = variable), alpha = 0.5) +
  facet_wrap(~variable, scales = "free_x") +
  theme_minimal() +
  labs(
    title = "Relationship Between Meteorological Variables and Burned Area",
    x = "Meteorological Variable Value",
    y = "Burned Area (hectares)"
  ) +
  scale_color_discrete(guide = "none")
```
