---
title: "Creating An Efficient Data Analysis Workflow"
format: markdown_github
---

This is a guided project for dataquest, investigating COVID-19 trends.

```{r}
#| include: false

data_file <- "book_reviews.csv"

if (!file.exists(data_file)) {
  download.file(
    "https://dq-content.s3.amazonaws.com/498/book_reviews.csv",
    destfile = data_file
  )
}
```



## 2. Getting Familiar with the Data

```{r}
library(readr)
library(dplyr)
```

Read the csv after downloading it and saving it in the working directory.

I specify `show_col_types = FALSE` to suppress the verbose messages showing
all of the column types it guessed. It's often useful to see this information
at least the first time you read in a dataset to confirm it is treating the 
columns as you would expect.

```{r}
book_reviews <- read_csv(
  "book_reviews.csv",
  show_col_types = FALSE
)

```

### 1.

```{r}
glimpse(book_reviews)
```

We see that there are `nrow(book_reviews)` rows and `ncol(book_reviews)` columns.

### 2. 

The column names are: `cat(colnames(book_reviews), sep = ", ")`.

They represent the book title, a one-word review, the state (whether this represents
where the review was written, or where the book was published, or something else is unknown), 
and the price.

### 3. 

The columns are all of type `character`, except for `price` which is `double`.

### 4.

```{r}
for (col in colnames(book_reviews)) {
  unique_vals <- unique(book_reviews[[col]])
  cat(paste0("Unique values in ", col, ":\n"))
  print(unique_vals)
}
```

## 3. Handling Missing Data

### 1. 

```{r}
for (col in colnames(book_reviews)) {
  missing_vals <- sum(is.na(book_reviews[[col]]))
  cat(paste0("Missing values in ", col, ": ", missing_vals, "\n"))
}
```

### 2.

```{r}
book_reviews_full <- book_reviews |>
  filter(!is.na(review))

dim(book_reviews_full)
```

### 3.

We removed `nrow(book_reviews) - nrow(book_reviews_full)` rows that had missing values in the `review` column.

### 4. 

First see the unique values so we know what we are working with:

```{r}
unique(book_reviews_full$state)
```

Then we can standardize:


```{r}
book_reviews_full <- book_reviews_full |>
  mutate(
    state = case_when(
      state == "NY" ~ "New York",
      state == "FL" ~ "Florida",
      state == "TX" ~ "Texas",
      state == "CA" ~ "California",
      .default = state
    )
  )

```

### 5. 


```{r}
book_reviews <- book_reviews |>
  mutate(
    review_num = case_when(
      review == "Poor" ~ 1,
      review == "Fair" ~ 2,
      review == "Good" ~ 3,
      review == "Great" ~ 4,
      review == "Excellent" ~ 5
    ),
    is_high_review = if_else(review_num >= 4, TRUE, FALSE)
  )
```

### 6. 

Calculate "Profitability":

```{r}
book_reviews |>
  group_by(book) |>
  summarize(
    n_purchased = n(),
    total_income = sum(price),
    avg_review = mean(review_num, na.rm = TRUE),
    n_reviews = sum(!is.na(review))
  )
```

"Fundamentals of R for Beginners" sold the most books, but "Secrets of R for Advanced Students"
had the highest total income. "Top 10 Mistakes R Beginners Make" has the highest average review.
