---
title: "Creating An Efficient Data Analysis Workflow"
format: markdown_github
---

This is a guided project for dataquest, investigating COVID-19 trends.

```{r}
#| include: false

data_file <- "book_reviews.csv"

if (!file.exists(data_file)) {
  download.file(
    "https://dq-content.s3.amazonaws.com/498/book_reviews.csv",
    destfile = data_file
  )
}
```



### 2. Getting Familiar with the Data

```{r}
library(tidyverse)
```

Read the csv after downloading it and saving it in the working directory.

I specify `show_col_types = FALSE` to suppress the verbose messages showing
all of the column types it guessed. It's often useful to see this information
at least the first time you read in a dataset to confirm it is treating the 
columns as you would expect.

```{r}
book_reviews <- read_csv(
  "book_reviews.csv",
  show_col_types = FALSE
)

```

#### 1.

```{r}
glimpse(book_reviews)
```

We see that there are `nrow(book_reviews)` rows and `ncol(book_reviews)` columns.

#### 2. 

The column names are: `cat(colnames(book_reviews), sep = ", ")`.

They represent the book title, a one-word review, the state (whether this represents
where the review was written, or where the book was published, or something else is unknown), 
and the price.

#### 3. 

The columns are all of type `character`, except for `price` which is `double`.

#### 4.

```{r}
for (col in colnames(book_reviews)) {
  # In each iteration of the loop, `col` gains a new value - the name of the 
  # column ("book", "review", "state", "price")
  # Use the `unique()` function to get the unique values of the current column
  unique_vals <- unique(book_reviews[[col]])
  # Print a heading for each iteration of the loop (\n means new line)
  cat(paste0("Unique values in ", col, ":\n"))
  # Then print the values
  print(unique_vals)
}
```

### 3. Handling Missing Data

#### 1. 

```{r}
for (col in colnames(book_reviews)) {
  missing_vals <- sum(is.na(book_reviews[[col]]))
  cat(paste0("Missing values in ", col, ": ", missing_vals, "\n"))
}
```

#### 2.

```{r}
book_reviews_full <- book_reviews |>
  filter(!is.na(review))

dim(book_reviews_full)
```

#### 3.

We removed `nrow(book_reviews) - nrow(book_reviews_full)` rows that had missing values in the `review` column.

#### 4. 

First see the unique values so we know what we are working with:

```{r}
unique(book_reviews_full$state)
```

Then we can standardize:


```{r}
book_reviews_full <- book_reviews_full |>
  mutate(
    state = case_when(
      state == "NY" ~ "New York",
      state == "FL" ~ "Florida",
      state == "TX" ~ "Texas",
      state == "CA" ~ "California",
      .default = state
    )
  )

```

#### 5. 


```{r}
book_reviews <- book_reviews |>
  mutate(
    review_num = case_when(
      review == "Poor" ~ 1,
      review == "Fair" ~ 2,
      review == "Good" ~ 3,
      review == "Great" ~ 4,
      review == "Excellent" ~ 5
    ),
    is_high_review = if_else(review_num >= 4, TRUE, FALSE)
  )
```

#### 6. 

Calculate "Profitability":

```{r}
book_reviews |>
  group_by(book) |>
  summarize(
    n_purchased = n(),
    total_income = sum(price),
    avg_review = mean(review_num, na.rm = TRUE),
    n_reviews = sum(!is.na(review))
  )
```

"Fundamentals of R for Beginners" sold the most books, but "Secrets of R for Advanced Students"
had the highest total income. "Top 10 Mistakes R Beginners Make" has the highest average review.

## Part 2 (Lesson 4; Specialized Data Processing)

### 1. Introduction

I downloaded the data from https://dsserver-prod-resources-1.s3.amazonaws.com/516/sales2019.csv and saved it in the working directory as `sales2019.csv`.

```{r}
library(tidyverse)

sales2019 <- read_csv("sales2019.csv", show_col_types = FALSE)
```

### 2. Data Exploration

Get the size (dimensions) and column names of the data:

```{r}
dim(sales2019)
names(sales2019)
```

We can get a prevew of the data and the data using `glimpse`:

```{r}
glimpse(sales2019)
```

We can see in the `glimpse` output that the `total_purchased` column contains some 
missing (`NA`) values, but we can check more comprehensively with:

```{r}
library(purrr)

# Check for any missing values in the entire data frame
anyNA(sales2019)

# Check for missing values in specific columns
anyNA(sales2019$title)

# Check for missing values in all columns
map(sales2019, anyNA)
```

We can also get a count of the number of missing values in each column:

```{r}
count_na <- function(x) {
  is_na <- is.na(x)
  sum(is_na)
}

count_na(sales2019$date)

count_na(sales2019$user_submitted_review)

map(sales2019, count_na)
```


It's also sometimes useful just to look at the data itself. You can click on the
data frame in the Environment pane in RStudio to view it in a spreadsheet-like format, 
or you can use the `View` function:

```{r eval=FALSE}
View(sales2019)
```

### 3. Handling Missing Data

#### Removing Rows with Missing Values in the `user_submitted_review` Column


```{r}
sales2019_cleaned <- sales2019 %>%
  filter(!is.na(user_submitted_review))

rows_removed <- nrow(sales2019) - nrow(sales2019_cleaned)
rows_removed
```

#### Fill in average for `total_purchased`

```{r}
average_purchased <- mean(sales2019_cleaned$total_purchased, na.rm = TRUE)

sales2019_cleaned <- sales2019_cleaned %>%
  mutate(
    total_purchased = if_else(
      is.na(total_purchased),
      average_purchased,
      total_purchased
    )
  )

# Check if there are any remaining missing values
anyNA(sales2019_cleaned$total_purchased)
```

#### Processing Review Data

Find positive reviews using `str_detect`:

Using `case_when()` is not necessary here, as we can directly create a logical column indicating whether the review is positive or not.

```{r}
library(stringr)

# This looks for the word "Awesome" in the user-submitted review
sales2019_cleaned <- sales2019_cleaned %>%
  mutate(
    positive_review = str_detect(
      user_submitted_review,
      "Awesome"
    )
  )

# We can also check if the review contains other positive words, check the reviews
# manually for examples.

# Enclosing the pattern in `str_detect` in `regex()` allows us to specify more
# options, such as case insensitivity.
sales2019_cleaned <- sales2019_cleaned %>%
  mutate(
    positive_review = str_detect(
      user_submitted_review,
      # strings separated by the pipe character `|` (or) are treated as alternatives
      regex("awesome|never read a better|learned a lot", ignore_case = TRUE)
    )
  )

```

#### Comparing Book Sales Between Pre- and Post-Program Sales

USe `?strftime` to see options for different ways of specifying the date format.

```{r}
sales2019_cleaned <- sales2019_cleaned %>%
  mutate(
    date = parse_date(date, format = "%m/%d/%y"),
  )
```

```{r}
program_start_date <- ymd("2019-07-01")
```

Create a new column indicating whether the sale was before or after the program started, and then summarize the total purchases and positive reviews before and after the program started.
Note that `positive_review` is a logical column, and when using mathematical functions on logical values, `TRUE` is interpreted as 1, and `FALSE` is interpreted as 0. 
Thus, summing `positive_review` gives the count of positive reviews, and calculating the mean gives the proportion of positive reviews.

```{r}
sales2019_cleaned <- sales2019_cleaned |>
  mutate(
    pre_post_program = if_else(
      date < program_start_date,
      "pre-progam",
      "post-program"
    )
  )

program_summary <- sales2019_cleaned |>
  group_by(pre_post_program) |>
  summarize(
    total_purchases = n(),
    total_books_purchased = sum(total_purchased, na.rm = TRUE)
  )

program_summary
```

#### Comparing Book Sales Within Customer Type

```{r}
program_summary_by_customer_type <- sales2019_cleaned |>
  group_by(pre_post_program, customer_type) |>
  summarize(
    total_purchases = n(),
    total_books_purchased = sum(total_purchased, na.rm = TRUE)
  )

program_summary_by_customer_type
```

#### Comparing Review Sentiment Between Pre- and Post-Program Sales

```{r}
program_summary_by_customer_type <- sales2019_cleaned |>
  group_by(pre_post_program, customer_type) |>
  summarize(
    total_purchases = n(),
    total_books_purchased = sum(total_purchased, na.rm = TRUE),
    positive_reviews = sum(positive_review, na.rm = TRUE),
    prop_positive_reviews = mean(positive_review, na.rm = TRUE)
  )

program_summary_by_customer_type
```
